### 2022-02-07 symposium.tex
Type and Null Safety for Regexes

% val date = new Regex("(\\d{4})-(\\d{2})-(\\d{2})")
% "2004-01-20" match {
%   case date(year, month, day) =>
%     s"$year was a good year for PLs."
% }

@@

Our contribution are as follows:

\begin{itemize}
  \item ...
  \item ...
  \item ...
  \item ...
\end{itemize}

@@

\section{Type-Level}

\section{Term-Level}

>>>>

\regexDocumentation
% val date = new Regex("(\\d{4})-(\\d{2})-(\\d{2})")
% "2004-01-20" match
%   case date(year, month, day) =>
%     s"$year was a good year for PLs."

@@

@@

We present our library in terms of 3 components:

\begin{enumerate}
  \item The type-level capture group analysis, which uses match types to inspect the user-provided regular expression and compute the corresponding tuple-based type representation (\Cref{sec:type-level}).

  \item The term-level capture group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed by the type-level component (\Cref{sec:term-level}).

  \item The user-level interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capture group extraction.
\end{enumerate}

We begin our presentation with the user-level interface, which takes the form of relatively simple Scala class and companion object:

\regexUserLevel
% object Regex:
%   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
%     new Regex(regex)
%
% class Regex[P] private (val regex: String):
%   val pattern = Pattern.compile(regex)
%   def unapply(s: String): Option[P] =
%     val m = pattern.matcher(s)
%     if (m.matches())
%       val a = Array.tabulate(m.groupCount) { i => m.group(i + 1) }
%       Some(toTuple[P](regex, a))
%     else
%       None

The |apply| method is the entry point of our library.
It takes a regular expression and returns an instance of |Regex|.
Instead of directly accepting |String| argument, that method requires a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression (the |Singleton| bound is necessary to guide Scala's type inference).
We use that type-level string immediately to instantiate |Compile|, which is our type-level capture group analysis component.
We store the output of that analysis as |Regex|'s type argument.

The implementation of the |Regex| class is straightforward.
Similarly its counterpart in the original Scala implementation, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.

The main difference between Scala's original |Regex| implementation and ours lies in the signature the |unapply| method, which determines the nature of extraction performed by that method.
By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capture groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
In our implementation, this is done via the |toTuple| method, which is our
term-level capture group processing component.
As a result, our implementation never returns null values, which removes the null-checking burden of from the user.

For instance, in the following example, we use a regular expression with an optional capture group to extract the integral and fractional parts of a rational number:

\regexRational
% val rational = Regex("(\\d+)\\.?(\\d+)?")
% "3.1415" match
%   case rational(i, Some(f)) =>
%     val n = i.size + f.size
%     s"This number is $n digits long"

\noindent
Here, our implementation instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which it then uses in |unapply|'s return type to obtain a precise representation of the regex's capture groups.

\section{Type-Level}
\label{sec:type-level}

The main challenge...

\section{Term-Level}
\label{sec:term-level}
